---
title: "HW5"
output: html_document
date: "2026-05-06"
---

```{r setup, include=FALSE}
library(tidyverse)
library(tidymodels)
library(e1071)
library(caret)
library(ISLR)
```

---

### Data setup
```{r}
auto = read_csv("auto.csv")
auto <- na.omit(auto)

auto$origin <- as.factor(auto$origin)
auto$mpg_cat <- as.factor(auto$mpg_cat)

set.seed(2) 

data_split <- initial_split(auto, prop = 0.7)
training_data <- training(data_split)
testing_data <- testing(data_split)
```

---

### 1-a. SVC (Linear Kernel)

```{r}
set.seed(2)
linear.tune <- tune.svm(mpg_cat ~ ., 
                        data = training_data, 
                        kernel = "linear", 
                        cost = exp(seq(-5, 2, len = 50))) 
plot(linear.tune)
best.linear <- linear.tune$best.model 
summary(best.linear)


# Training Error Rate
pred.train.lin <- predict(best.linear, newdata = training_data)
train_err_lin <- mean(pred.train.lin != training_data$mpg_cat)

# Test Error Rate
pred.test.lin <- predict(best.linear, newdata = testing_data) 
test_err_lin <- mean(pred.test.lin != testing_data$mpg_cat)

# Results
cat("SVC Training Error:", train_err_lin, "\n")
cat("SVC Test Error:", test_err_lin, "\n")
```

Training Error = 0.0912
Test Error = 0.0508

---

### 1-b. SVM (Radial Kernel)

```{r}
# Fit and tune SVM with Radial Kernel
set.seed(2)
radial.tune <- tune.svm(mpg_cat ~ ., 
                        data = training_data, 
                        kernel = "radial", 
                        cost = exp(seq(1, 7, len = 50)), 
                        gamma = exp(seq(-10, -2, len = 20))) 

plot(radial.tune, transform.y = log, transform.x = log,
color.palette = terrain.colors)

best.radial <- radial.tune$best.model 
summary(best.radial)


# Training Error Rate
pred.train.rad <- predict(best.radial, newdata = training_data)
train_err_rad <- mean(pred.train.rad != training_data$mpg_cat)

# Test Error Rate
pred.test.rad <- predict(best.radial, newdata = testing_data) 
test_err_rad <- mean(pred.test.rad != testing_data$mpg_cat)

# Results
cat("SVM Training Error:", train_err_rad, "\n")
cat("SVM Test Error:", test_err_rad, "\n")
```

Training Error = 0.0912
Test Error = 0.0508

---

### 1-c. Boosting & Comparison with SVC and SVM

```{r}
ctrl <- trainControl(method = "cv")

# SVC (Linear Kernel)
set.seed(2) 
svml.fit <- train(mpg_cat ~ ., 
                  data = training_data, 
                  method = "svmLinear2", 
                  tuneGrid = data.frame(cost = linear.tune$best.parameters$cost), 
                  trControl = ctrl) 

# SVM (Radial Kernel)
set.seed(2)
svmr.fit <- train(mpg_cat ~ ., 
                  data = training_data, 
                  method = "svmRadial", 
                  tuneGrid = data.frame(C = radial.tune$best.parameters$cost,
                                        sigma = radial.tune$best.parameters$gamma),
                  trControl = ctrl) 

# Boosting
set.seed(2)
boost.fit <- train(mpg_cat ~ ., 
                   data = training_data, 
                   method = "gbm", 
                   trControl = ctrl, 
                   verbose = FALSE) 

# Comparison
resamp <- resamples(list(SVC = svml.fit, 
                         SVM_radial = svmr.fit,
                         Boosting = boost.fit))

summary(resamp)
bwplot(resamp)
```

Boosting outperformed both SVC (Linear Kernel) and SVM (Radial Kernel) models with the highest mean Kappa statistic (0.839) and mean Accuracy (0.920).
However, the mean accuracy for the SVM Radial Kernel model (0.905) was higher than the SVC Linear Kernel mode (0.898). The Support Vector Machine with a radial kernel also outperformed the linear support vector classifier in the Kappa statistic suggesting that the radial model is more robust.
These results indicate that the Support Vector Machine with a radial kernel provides a better predictive performance on the data than the linear support vector classifier but that boosting outperformed both.

---

### 2-a. Hierarchical Clutering

```{r}
# Hierarchical Clustering with Complete Linkage
states_data <- USArrests
hc_complete <- hclust(dist(states_data), method = "complete")

clusters <- cutree(hc_complete, k = 3)

sort(clusters)
table(clusters)

# dendrogram
plot(hc_complete, main = "Dendrogram of USArrests (Unscaled)", xlab = "States", sub = "")

rect.hclust(hc_complete, k = 3, border = "red")
```

This dendrogram visuzalizes the states that make up each cluster. 16 states were in cluster 1, 14 in cluster 2 and 20 in cluster 3. 

Cluster 1: Alabama, Alaska, Arizona, California, Delaware, Florida, Illinois, Louisiana, Maryland, Michigan, Mississippi, Nevada, New Mexico, New York, North Carolina, South Carolina.  


Cluster 2: Arkansas, Colorado, Georgia, Massachusetts, Missouri, New Jersey, Oklahoma, Oregon, Rhode Island, Tennessee, Texas, Virginia, Washington, Wyoming.  


Cluster 3: Connecticut, Hawaii, Idaho, Indiana, Iowa, Kansas, Kentucky, Maine, Minnesota, Montana, Nebraska, New Hampshire, North Dakota, Ohio, Pennsylvania, South Dakota, Utah, Vermont, West Virginia, Wisconsin.




### 2-b. Scaling
```{r}
# Scaling (standard deviation = 1)
states_scaled <- scale(USArrests)

# Hierarchical clustering with complete linkage and Euclidean distance
hc_scaled <- hclust(dist(states_scaled), method = "complete")

clusters_scaled <- cutree(hc_scaled, k = 3)

table(clusters_scaled)

# Comparing
table(Unscaled = clusters, Scaled = clusters_scaled)

# Scaled data dendrogram
plot(hc_scaled, main = "Hierarchical Clustering: Scaled Data (SD = 1)", xlab = "", sub = "") 
rect.hclust(hc_scaled, k = 3, border = "blue")
```

Scaling the variables significantly changed the clustering results. The table shows that only 6 of the original 16 states in the unscaled clusters remained with 9 moving to cluster 2 and 1 moving to cluster 3. Out of the original 14 in the unscaled cluster 2, 2 were shifted to cluster 1 and 10 were shifted to cluster 3. The original states in cluster 3 all remained, indicating that it was the most stable of the three unscaled clusters. The scaled data yielded 8 states in cluster 1, 11 states in cluster 2 and 31 states in cluster 3. The dendrogram visualizes the scaled data clusters to highlight the differences. 

In my opinion, the data should absolutely be scaled prior to inter-observation computation of dissimilarities because otherwise the variables on a small scale like murder (~10^1 scale) get largely ignored by the model as the variables on a larger scale like assault (~10^2 scale) over power the results. Scaling the data puts the clustering on an even field. 
